Gradual Soundness: Lessons from Static Python

### 2022-02-01 pj-7.1/paper.tex
pj: kc / grammarly fixes
oh dear 'fuction' how can I mis-spell that

Gradual typing has attracted significant interest as a solution to
the debate between static and dynamic typing.
The premise is simple: let programmers introduce types in part of a
codebase and while leaving the rest untyped.
Run-time casts and checks enforce the assumptions that typed code makes about
untyped components, thereby ensuring that the types are sound no matter how
untyped code behaves.

@@

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes.
    At Instagram, the web server has become significantly more efficient
    after the application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    \numbenchmark{} public microbenchmarks~(\cref{s:microbenchmarks}).
    Using only shallow types, \SP{} runs slightly faster than Python
    despite the costs of enforcing soundness~(avg. \pctnum{28}).
    With fine-tuned types, \SP{} consistently out-performs the Python baseline~(avg. \pctnum{67}).

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design.
In particular, the maintainers of optionally-typed languages
may wish to focus on shallow types and JIT compilation as a
first step toward sound gradual types.
Second, we want to entice researchers.
The \SP{} type system has many restrictions that are suprising
at first glance.
For example, functions are supported only by shallow types
and method overrides that use the dynamic type are more constrained
than overrides in untyped code.
Some of these restrictions might be lifted by future research.
Others might be useful to adopt in new language designs.

@@

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is provide a second kind of dictionary that ensures
the validity of its elements.
These different values have different types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a constant-time,
    compound tag check for concrete dict values that were initialized with
    exactly the same key and value types.
    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.

@@

The upside of concrete types is that they enforce stronger type constraints.
These constraints can catch bugs and always lead to faster typed code.
Faster performance is not guaranteed in general, though, because concrete types must
check all writes from untyped contexts.
For comparison, \cref{t:shallow-vs-concrete} presents the worst-case costs of
shallow and concrete dict types.
Shallow types pay a constant-time cost whenever an untyped value enters typed code
and whenever a typed context reads from a shallow value.
Concrete types pay a potentially-high cost for constructors,
a constant-time cost for casts and for writes in untyped code,
and zero cost for reads.
If a value travels to untyped code, gets written to many times, and gets read few times,
then the shallow types may run faster in the end.

@@

In Python, the syntax for expressions overlaps with the syntax of types.
For example, the Python name \code{None} is both an expression and a type,
and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
\Cref{f:surface-types} does not follow these standards; instead it keeps the two syntaxes distinct.
Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
names (\code{None}, \code{CheckedDict}).

@@

\SP{} evaluation types can all be enforced fully and efficiently
with casts.
\Cref{t:cast} describes the casts in detail.
Most call for a tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Even checked dictionaries rely on tag checks.
A checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, this exact-match rules applies even when $\spteval_0$
or $\spteval_1$ is the dynamic type.

@@

The first two rules are for function application.
A typed function may be applied to an argument that matches its domain type,
in which case it computes a value that matches its codomain type.
A dynamically-typed variable may be applied to any input (that matches the top
type $\sptobject$) to yield a dynamically-typed result.
The next rules illustrate writes to Python dictionaries and checked dictionaries.
A shallow Python dict may be updated with any kind of key and value.
By contrast, a concrete checked dict requires keys and values that match its type parameters.
Next we have two rules for classes:
\trule{F-Set} says that writes to a class field must match the declared field
type; and \trule{M-App} shows that typed methods impose similar constraints as
typed fuctions.

@@

\SP{} does not attempt to type first-class classes: partly because they have
yet to appear in performance-critical code in the Instagram web server
and partly because they do not fit well with a nominal type system.
The straightforward but restrictive approach for first-class nominal class types
is to force code that uses a first-class class to expect subtypes of a
particular named static class.
Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
structural objects.

@@

To measure the effect of a code change on these hosts, an internal profiling tool selects
two representative sets of machines to run as experimental and control groups.
The experimental groups gets the latest version of the server code;
the control group gets the previous version.
Next, the profiling tool slowly increases the number of requests that these
machines receive until each is running at maximum load.
Once the machines are fully allocated, the tool measures how many requests
per second each group is able to handle.
If the experimental group can serve more requests per second, then the change
is a success.
Assuming that the experimental and control groups have equivalent hardware
capabilities, and assuming that both groups receive a representative sample
of requests, then the results of an experiment (increase/decrease in CPU efficiency)
should predict actual performance.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython improved performance by compiling these modules to C,
but its partial adoption had led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

@@

\begin{enumerate}
  \item
    \ipara{Use primitive types}
    Integers and booleans in hot code paths run fastest with
    primitive type~(\cref{s:c-types}).
    %%As noted above, primitives were required for \SP{} to match the performance of
    %%Cython.
    Twelve modules currently use primitives.
    %% - bg: focus on block, convert, no problem?

@@

  \item
    \ipara{Convert \code{@classmethod} to \code{@staticmethod}}
    Static methods can get invoked directly as functions, bypassing the class vtable.
    Twenty-four modules use static methods, though many of these uses pre-date \SP{}.
    %% 2022-01-12 Carl:
    %% I don’t think it was a super common change, but there were some cases where we
    %% changed a @classmethod that didn’t actually use its cls argument to a
    %% @staticmethod. It looks like we have 24 static modules using a
    %% @staticmethod somewhere in the module, but I think in the majority of those
    %% there just happened to be some pre-existing staticmethods (they aren’t
    %% uncommon in our codebase); I think a minority were intentional
    %% optimizations on our part. And the pre-existing ones may or may not be
    %% meaningfully improving performance. So I’m not sure how valuable it is to
    %% report this metric.

@@

In conclusion, the \SP{} approach to gradual typing seems to be a promising
way to realize the vision of gradual typing.
Determining whether this conjecture holds more broadly
calls for research on two fronts.
First, \SP{} must be applied to additional projects in a module-at-a-time manner.
A careful evaluation~\cite{gtnffvf-jfp-2019} of a few small projects would be
especially useful to find performance bottlenecks.
Second, the \SP{} approach should be adapted to other languages.
An optionally language such as TypeScript would be an ideal starting
point.
%% Gradual, progressive soundness may prove itself to be an effective general strategy for
%% gradual types in practice.

>>>>

Gradual typing has attracted significant interest as a solution to
the debate between static and dynamic typing.
The premise is simple: let programmers introduce types in part of a
codebase while leaving the rest untyped.
Run-time casts and checks enforce the assumptions that typed code makes about
untyped components, thereby ensuring that the types are sound no matter how
untyped code behaves.

@@

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes.
    At Instagram, the web server has become significantly more efficient
    after the application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    \numbenchmark{} public microbenchmarks~(\cref{s:microbenchmarks}).
    Using only shallow types, \SP{} runs slightly faster than Python
    despite the costs of enforcing soundness~(avg. \pctnum{28}).
    With fine-tuned types, \SP{} consistently outperforms the Python baseline~(avg. \pctnum{67}).

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design.
In particular, the maintainers of optionally-typed languages
may wish to focus on shallow types and JIT compilation as a
first step toward sound gradual types.
Second, we want to entice researchers.
The \SP{} type system has many noteworthy restrictions.
For example, functions are supported only by shallow types
and method overrides that use the dynamic type are more constrained
than overrides in untyped code.
Some of these restrictions might be lifted by future research.
Others might be useful to adopt in new language designs.

@@

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is to provide a second kind of dictionary that ensures
the validity of its elements.
These different values have different types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a constant-time,
    compound tag check for concrete dict values that were initialized with
    exactly the same key and value types.
    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.

@@

The upside of concrete types is that they enforce stronger type constraints.
These constraints can catch bugs and always lead to faster typed code.
Faster performance is not guaranteed in general, though, because concrete types must
check all writes from untyped contexts.
For comparison, \cref{t:shallow-vs-concrete} presents the worst-case costs of
shallow and concrete dict types.
Shallow types pay a constant-time cost whenever an untyped value enters typed code
and whenever a typed context reads from a shallow value.
Concrete types pay a potentially-high cost for constructors,
a constant-time cost for casts and for writes in untyped code,
and zero cost for reads.
If a value travels to untyped codes and receives many writes but few reads,
then the shallow types may run faster in the end.

@@

In Python, the syntax for expressions overlaps with the syntax of types.
For example, the Python name \code{None} is both an expression and a type,
and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
\Cref{f:surface-types} does not follow these standards; instead it keeps the two syntaxes distinct.
Expressions use lowercase names (\code{none}, \code{chkdict}) and types use capitalized
names (\code{None}, \code{CheckedDict}).

@@

\SP{} evaluation types can all be enforced fully and efficiently
with casts.
\Cref{t:cast} describes the casts in detail.
Most call for a tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Even checked dictionaries rely on tag checks.
A checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, this exact-match rule applies even when $\spteval_0$
or $\spteval_1$ is the dynamic type.

@@

The first two rules are for function application.
A typed function may be applied to an argument that matches its domain type,
in which case it computes a value that matches its codomain type.
A dynamically-typed variable may be applied to any input (that matches the top
type $\sptobject$) to yield a dynamically-typed result.
The next rules illustrate writes to Python dictionaries and checked dictionaries.
A shallow Python dict may be updated with any kind of key and value.
By contrast, a concrete checked dict requires keys and values that match its type parameters.
Next we have two rules for classes:
\trule{F-Set} says that writes to a class field must match the declared field
type; and \trule{M-App} shows that typed methods impose similar constraints as
typed functions.

@@

\SP{} does not attempt to type first-class classes: partly because they have
yet to appear in performance-critical code in the Instagram web server
and partly because they do not fit well with a nominal type system.
The straightforward but restrictive approach for first-class nominal class types
is to force code that uses a first-class class to expect subtypes of a
particular named static class.
Flatt et al.~\cite{fkf-popl-1998} propose a more flexible approach, %(specifically for mixins)
but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
structural objects.

@@

To measure the effect of a code change on these hosts, an internal profiling tool selects
two representative sets of machines to run as experimental and control groups.
The experimental group gets the latest version of the server code;
the control group gets the previous version.
Next, the profiling tool slowly increases the number of requests that these
machines receive until each is running at maximum load.
Once the machines are fully allocated, the tool measures how many requests
per second each group is able to handle.
If the experimental group can serve more requests per second, then the change
is a success.
Assuming that the experimental and control groups have equivalent hardware
capabilities, and assuming that both groups receive a representative sample
of requests, then the results of an experiment (increase/decrease in CPU efficiency)
should predict actual performance.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython improved performance by compiling these modules to C,
but its partial adoption had led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} lets programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

@@

\begin{enumerate}
  \item
    \ipara{Use primitive types}
    Integers and booleans in hot code paths run fastest with
    primitive types~(\cref{s:c-types}).
    %%As noted above, primitives were required for \SP{} to match the performance of
    %%Cython.
    Twelve modules currently use primitives.
    %% - bg: focus on block, convert, no problem?

@@

  \item
    \ipara{Convert \code{@classmethod} to \code{@staticmethod}}
    Static methods can get invoked directly as functions, bypassing the class vtable.
    Twenty-four modules use static methods (many of these pre-date \SP{}).
    %% 2022-01-12 Carl:
    %% I don’t think it was a super common change, but there were some cases where we
    %% changed a @classmethod that didn’t actually use its cls argument to a
    %% @staticmethod. It looks like we have 24 static modules using a
    %% @staticmethod somewhere in the module, but I think in the majority of those
    %% there just happened to be some pre-existing staticmethods (they aren’t
    %% uncommon in our codebase); I think a minority were intentional
    %% optimizations on our part. And the pre-existing ones may or may not be
    %% meaningfully improving performance. So I’m not sure how valuable it is to
    %% report this metric.

@@

In conclusion, the \SP{} approach to gradual typing seems to be a promising
way to realize the vision of gradual typing.
Determining whether this conjecture holds more broadly
calls for research on two fronts.
First, \SP{} must be applied to additional projects in a module-at-a-time manner.
A careful evaluation~\cite{gtnffvf-jfp-2019} of a few small projects would be
especially useful to find performance bottlenecks.
Second, the \SP{} approach should be adapted to other languages.
An optionally-typed language such as TypeScript would be an ideal starting
point.
%% Gradual, progressive soundness may prove itself to be an effective general strategy for
%% gradual types in practice.
